---
title: "final_project"
author: "Eden Doron, Tomer Goldstein"
date: "8/20/2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we will load the data from the paper's dataset and organize the matrix to be with clean data only.
```{r loading data}
data <- read.csv('./data/E-MTAB-2758-raw-counts.tsv', header = T, sep = '\t', row.names = 'Gene.ID')

data_matrix <- as.matrix(data)
gene_map <- data_matrix[ ,c(1)]
raw_data <- data_matrix[, -1]

experiment_design_data <- read.csv('./data/E-MTAB-2758-experiment-design.tsv', header = T, sep = '\t', row.names = 1, stringsAsFactors = T)
experiment_design <- experiment_design_data[, c(11, 13)]
colnames(experiment_design) <- c('Time', 'Control')
```

Sort column names according to sample number
```{r sort data}
raw_data <- apply(raw_data, MARGIN=c(1,2), FUN=as.integer)
raw_data_colnames <- colnames(raw_data)
sorted_raw_data <- stringr::str_sort(raw_data_colnames)
raw_data <- raw_data[, sorted_raw_data]
```

Next, we will perform differential expression on the sorted samples and normalize it as seen in the tutorial
```{r differential expression}
library(DESeq2)

dds <- DESeqDataSetFromMatrix(countData = raw_data,
                              colData = experiment_design,
                              design = ~ Time + Control)
table <- head(assay(dds))

dds <- dds[rowSums(counts(dds)) > 1,]

vsd <- vst(dds, blind = FALSE)
vsd_raw_data <- vst(raw_data, blind = FALSE)
rownames(vsd_raw_data) <- gene_map[rownames(vsd_raw_data)]
dds_analysis <- DESeq(dds)
res <- results(dds_analysis)

vsd_subset <- vsd_raw_data[c("GPR84","SERPINE1","PDGFB","NIPAL4","LIF", "FOSL1"),]
scaled_vsd_subset <- t(scale(t(vsd_subset)))
```

Presenting heatmap after calculating the distances between samples
```{r}
library ("pheatmap")
library ("RColorBrewer")

sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste(vsd$Control, vsd$Time, sep = " - ")
colors <- colorRampPalette( rev(brewer.pal(9, "Reds")) )(255)

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```

Heatmap with top 6 genes
```{r}
anno <- as.data.frame(colData(vsd)[, c("Control","Time")])
# removing 0 hour samples
anno <- anno[-c(1,2,3,4,5,6), ]
scaled_vsd_subset <- scaled_vsd_subset[, -c(1,2,3,4,5,6)]
pheatmap(scaled_vsd_subset, annotation_col = anno)
```

Conducting a DESeq analysis and filtering the results by a log2 fold change threshold of 1 as well as taking the most up-regulated and down-regulated genes with a adjusted p-value smaller than 0.1.
```{r }
library("EnhancedVolcano")
EnhancedVolcano(res, lab = gene_map[rownames(res)], 
                x = 'log2FoldChange', y = 'pvalue',
                selectLab = c("GPR84", "SERPINE1", "PDGFB", "LIF", "NIPAL4", "FOSL1"),
                title = 'Volcano Plot',
                labSize = 6, pCutoff = 10e-20, FCcutoff = 1)
```

```{r}
res <- results(dds_analysis, lfcThreshold = 1)
summary(res)


resSig <- subset(res, padj < 0.1)
down_regulated <- head(resSig[ order(resSig$log2FoldChange), ])
rownames(down_regulated) <- gene_map[rownames(down_regulated)]
down_regulated <- data.frame(down_regulated)

up_regulated <- head(resSig[order(resSig$log2FoldChange, decreasing = TRUE), ])
rownames(up_regulated) <- gene_map[rownames(up_regulated)]
up_regulated <- data.frame(up_regulated)

```

Clustring
```{r}
library("factoextra")
library("Rtsne")
set.seed(42)

experiment_design <- experiment_design_data[, c(1, 11, 13)]
colnames(experiment_design) <- c('Sample', 'Time', 'Control')

raw_data_clean <- raw_data[rowSums(raw_data[, -1]) > 0 , ]
scaled_data <- scale(na.omit(t(raw_data_clean)))

fviz_nbclust(scaled_data, kmeans, method = "silhouette")
fviz_nbclust(scaled_data, kmeans, method = "wss")

# kmeans
km_result2 <- kmeans(scaled_data, centers = 2, nstart = 25)
km_result5 <- kmeans(scaled_data, centers = 5, nstart = 25)

tsne <- Rtsne(t(raw_data), perplexity = 5)

plot(tsne$Y, col = km_result2$cluster)
plot(tsne$Y, col = km_result5$cluster)

# hclust
hc_result <- hclust(dist(scaled_data), method = "ward.D")
hc_clusters <- cutree(hc_result, k = 2)
plot(tsne$Y, col = hc_clusters)

hc_result <- hclust(dist(scaled_data), method = "ward.D")
hc_clusters <- cutree(hc_result, k = 5)
plot(tsne$Y, col = hc_clusters)

# manual clustring by labels
clusters_by_treatment <- as.vector(experiment_design$Control)
clusters_by_treatment[clusters_by_treatment == "control"] <- 1
clusters_by_treatment[clusters_by_treatment == "osmium_based_compound"] <- 2
clusters_by_treatment <- as.integer(clusters_by_treatment)
names(clusters_by_treatment) <- experiment_design$Sample

plot(tsne$Y, col = clusters_by_treatment)

clusters_by_time <- as.vector(experiment_design$Time)
clusters_by_time[clusters_by_time == "0_hour"] <- 1
clusters_by_time[clusters_by_time == "4_hour"] <- 2
clusters_by_time[clusters_by_time == "12_hour"] <- 3
clusters_by_time[clusters_by_time == "24_hour"] <- 4
clusters_by_time[clusters_by_time == "48_hour"] <- 5
clusters_by_time <- as.integer(clusters_by_time)
names(clusters_by_time) <- experiment_design$Sample

plot(tsne$Y, col = clusters_by_time)
```

